\documentclass[12pt]{article}

%% ---------------------------------------------------------------
%% Packages
%% ---------------------------------------------------------------
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{enumitem}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{natbib}

%% ---------------------------------------------------------------
%% Theorem environments
%% ---------------------------------------------------------------
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}
\newtheorem{observation}{Observation}

%% ---------------------------------------------------------------
%% Custom commands
%% ---------------------------------------------------------------
\newcommand{\Qm}{Q_m}
\newcommand{\Pm}{P_m}
\newcommand{\dH}{d_H}
\newcommand{\Bm}{B(m,r)}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}

%% ---------------------------------------------------------------
%% Metadata
%% ---------------------------------------------------------------
\title{%
  Primes in the Odd Binary Hypercube:\\
  Search Geometry, Distribution Geometry, and Sieve Repulsion at the Edge Scale
}

\author{
  Preetam Ojha\\
  \texttt{nehu.preet@gmail.com}
}

\date{May 2026}

%% arXiv subject classification
%% Primary:   cs.DM  (Discrete Mathematics)
%% Secondary: cs.DS  (Data Structures and Algorithms)
%% Suggested cross-list: math.CO, math.NT

%% ---------------------------------------------------------------
\begin{document}
%% ---------------------------------------------------------------

\maketitle

\begin{center}
\small
\textbf{Suggested arXiv subject classes:} cs.DM (primary), cs.DS (secondary);
possible cross-list: math.CO, math.NT
\end{center}

\begin{abstract}
We embed prime numbers as vertices in the $m$-bit odd binary hypercube
$Q_m^{\mathrm{odd}} = \{x\in\{0,1\}^m : \mathrm{bit}_0(x)=1\}$, where
distance is measured by Hamming weight rather than arithmetic difference.
Two complementary experimental questions are studied: (i)~\emph{search
geometry}---how does Hamming-ball prime search compare with standard
odd-increment search?---and (ii)~\emph{distribution geometry}---do primes
behave like a random subset of the odd hypercube?

On search geometry, we find that standard odd-increment search (SOI) achieves
the lowest average candidate-check count, but low-bit-first Hamming search
(LBF) wins or ties SOI on 58--63\% of individual starting points at
$m\in\{24,28,32\}$.  Pareto analysis confirms that the two families optimize
different objectives: SOI finds arithmetically close primes, while Hamming
search finds bit-close primes.

On distribution geometry, we report two contrasting findings.
\emph{Hamming-ball pseudorandomness:} prime counts inside Hamming balls of
radius 1--4 fluctuate with z-scores close to $N(0,1)$, statistically
indistinguishable from a random subset of the same density.  Low-degree
Walsh--Fourier coefficients of the centered prime indicator are similarly
small, consistent with prime pseudorandomness at coarse geometric and algebraic
scales.
\emph{Hamming-edge sieve repulsion:} at the single-edge scale, prime-prime
Hamming-1 edge counts are approximately one-third below the random-model
prediction, with ratio $\rho\approx 0.67$ stable across $m=10$--$22$ and
uniform across all bit positions.  This is consistent with the
Hardy--Littlewood twin-prime sieve correction factor $C_2\approx 0.66$.

The headline finding is a clean contrast:
\emph{primes look random inside Hamming balls, but repel along hypercube edges}.
This distinction---pseudorandomness at coarse scales, sieve repulsion at the
edge scale---is a new experimental observation and is not obvious at the outset.
A parallel distinction separates search experiments from distribution
experiments: bit-position bias observed in search paths does not appear in the
prime-prime edge geometry, where the edge deficit is uniform across bit
directions.

No new theorems are proved; all claims are empirical.

We release all code at \url{https://github.com/pojhafb/prime-hypercube-search}.
\end{abstract}

\tableofcontents
\newpage

%% ---------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
%% ---------------------------------------------------------------

Prime numbers are classically studied as points on the integer number line,
where distance is measured by subtraction: $|p-x|$. The prime number theorem
\citep{hadamard1896, delavalleepoussin1896} tells us that primes are sparse:
among integers near $N$, about $1/\ln N$ of them are prime. The expected gap to
the next prime from a typical integer near $x$ is approximately $\ln x$.

Every integer also has a binary representation, and binary strings of length
$m$ are naturally placed as vertices of the $m$-dimensional hypercube
$Q_m = \{0,1\}^m$. Two vertices are adjacent if they differ in exactly one bit.
The number of differing bit positions between two integers $x$ and $y$ is the
\emph{Hamming distance}:
\[
  \dH(x,y) = |\{i : \mathrm{bit}_i(x) \neq \mathrm{bit}_i(y)\}|.
\]
This defines a metric on $Q_m$ that is very different from arithmetic distance.

This paper asks: \emph{what does the distribution of primes look like in
Hamming geometry?} More concretely, can one find a prime near a given integer
$x$ by searching within a small Hamming ball around $x$, and how does such
search compare to the standard algorithm of testing $x, x+2, x+4, \ldots$?

\subsection{Why this question is interesting}

Two independent motivations drive this work.

\paragraph{Search geometry.}
Finding a prime near a given large integer is a basic primitive in
cryptographic key generation \citep{menezes1996}. Standard algorithms test
$x, x+2, x+4, \ldots$ and rely on the prime number theorem for their expected
cost. An alternative search in Hamming space produces primes with different
characteristics---primes that may be arithmetically far but differ in only a
few bits. The goal is not to beat standard search on average, but to
characterize when and how the two search geometries differ.

\paragraph{Distribution geometry.}
The primes are a fixed arithmetically-defined set. Placing them as a subset of
the hypercube $Q_m$ with Hamming distance and asking whether they look
``random'' at various scales is a natural question in discrete mathematics.
It connects to the Boolean Fourier analysis of $\{0,1\}^m$
\citep{odonnell2014}, and the answer turns out to depend strongly on scale:
pseudorandom at the Hamming-ball level, non-random at the single-edge level.
This scale-dependent picture is not predicted by any simple argument and
constitutes the main empirical discovery of the paper.

\subsection{Summary of results}

\paragraph{Search geometry.}

\begin{enumerate}[label=(\roman*)]
  \item \textbf{Hamming density (empirical).} In a full scan of $Q_{14}$,
    98.4\% of integers are within Hamming distance 2 of a prime and 100\%
    within distance 3.

  \item \textbf{Parity dominates the full hypercube.} In unrestricted Hamming
    search, bit~0 dominates because flipping it toggles parity. This motivates
    the odd-subcube formulation.

  \item \textbf{SOI wins average checks; LBF is competitive per-instance.}
    Standard odd-increment search achieves the lowest average candidate-check
    count. However, low-bit-first Hamming search wins or ties SOI on 58--63\%
    of individual starting points at $m\in\{24,28,32\}$.

  \item \textbf{Two search geometries.} Standard search finds arithmetically
    close primes; Hamming search finds bit-close primes. Pareto analysis shows
    neither family dominates the other when both objectives are considered.
\end{enumerate}

\paragraph{Distribution geometry.}

\begin{enumerate}[label=(\roman*)]
  \setcounter{enumi}{4}
  \item \textbf{Hamming-ball pseudorandomness.} Prime counts inside Hamming
    balls of radius 1--4 fluctuate with normalized z-scores close to $N(0,1)$,
    matching a random subset of the same density. Low-degree Walsh--Fourier
    coefficients of the centered prime indicator are similarly small.

  \item \textbf{Hamming-edge sieve repulsion.} Prime-prime Hamming-1 edge
    counts are $\approx$one-third below the random-model expectation,
    $\rho\approx 0.67$ across $m=10$--$22$.  The deficit is uniform across all
    bit directions.  This is consistent with the Hardy--Littlewood sieve
    correction for prime pairs.

  \item \textbf{Two phenomena are distinct.} Bit-position bias (low bits
    preferred in search paths) and Hamming-1 edge repulsion (uniform across bit
    directions) are separate effects.  Search-path bias and prime-prime edge
    geometry probe different aspects of prime distribution.
\end{enumerate}

\subsection{Organization}

Section~\ref{sec:model} sets up the mathematical model.
Section~\ref{sec:strategies} describes the search strategies.
Section~\ref{sec:theory} provides heuristic estimates.
Section~\ref{sec:experiments} presents search-geometry experiments.
Section~\ref{sec:pareto} gives the Pareto analysis.
Section~\ref{sec:distribution} presents distribution-geometry experiments
(Hamming-ball discrepancy, Fourier analysis, and Hamming-1 edge counts).
Section~\ref{sec:related} discusses related work.
Section~\ref{sec:conclusion} concludes.

%% ---------------------------------------------------------------
\section{Mathematical Model}
\label{sec:model}
%% ---------------------------------------------------------------

\subsection{The binary hypercube}

\begin{definition}
  The \emph{$m$-bit binary hypercube} is the graph
  $Q_m=(\{0,1\}^m,E)$ where $\{x,y\}\in E$ if and only if
  $\dH(x,y)=1$.
\end{definition}

We identify each vertex with the integer it encodes in binary, so $Q_m$
contains the integers $\{0,1,\ldots,2^m-1\}$.

\begin{definition}
  The \emph{Hamming ball} of radius $r$ centered at $x$ is:
  \[
    \mathcal{B}(x,r)=\{y\in Q_m : \dH(x,y)\leq r\}.
  \]
  Its size depends only on $m$ and $r$:
  \[
    |\mathcal{B}(x,r)| = B(m,r)=\sum_{i=0}^{r}\binom{m}{i}.
  \]
\end{definition}

\subsection{Primes in the hypercube}

\begin{definition}
  The \emph{prime set} of $Q_m$ is
  \[
    P_m=\{x\in Q_m : x \text{ is prime}\}.
  \]
  The \emph{Hamming distance to the nearest prime} from $x$ is:
  \[
    D_m(x)=\min_{p\in P_m}\dH(x,p).
  \]
\end{definition}

By the prime number theorem, $|P_m| \approx 2^m/(m\ln 2)$, so the prime density
in $Q_m$ is approximately:
\[
  \delta_m \approx \frac{1}{m\ln 2}.
\]

\subsection{The odd subcube}

All primes greater than 2 are odd. Therefore, for $m\geq 2$, all nontrivial
prime vertices lie in:
\[
  Q_m^{\mathrm{odd}}
  =
  \{x\in Q_m : \mathrm{bit}_0(x)=1\}.
\]
$Q_m^{\mathrm{odd}}$ is the odd subcube, a sub-hypercube of dimension $m-1$
with $2^{m-1}$ vertices. In this paper, odd-subcube Hamming search fixes bit~0
to 1 and flips only bits $1,\ldots,m-1$.

%% ---------------------------------------------------------------
\section{Search Strategies}
\label{sec:strategies}
%% ---------------------------------------------------------------

Given a starting integer $x$, we set:
\[
  x_0 = x\,|\,1,
\]
that is, $x_0$ is the odd integer obtained by setting bit~0 to 1. A search
strategy returns a prime $p$, a candidate-check count $k$, the Hamming distance
$\dH(x_0,p)$, and the arithmetic distance $|p-x_0|$.

We compare six strategies.

\subsection{Standard odd-increment search (SOI)}

\begin{algorithm}[H]
\caption{Standard odd-increment search}
\begin{algorithmic}[1]
\State $x_0 \gets x \,|\, 1$
\State $k \gets 0$, $c \gets x_0$
\Loop
  \State $k \gets k + 1$
  \If{$c$ is prime}
    \Return $(c,\, k,\, \dH(x_0,c),\, c-x_0)$
  \EndIf
  \State $c \gets c+2$
\EndLoop
\end{algorithmic}
\end{algorithm}

This is the standard arithmetic-local algorithm. It always finds the smallest
prime $\geq x_0$.

\subsection{Odd-subcube Hamming search}

Let $\mathcal{F}(m,r_{\max})$ be a list of flip sets---tuples of bit positions
in $\{1,\ldots,m-1\}$---ordered by some policy $\pi$.

\begin{algorithm}[H]
\caption{Odd-subcube Hamming search}
\begin{algorithmic}[1]
\State $x_0 \gets x \,|\, 1$
\State $k \gets 0$
\For{each flip set $S \in \mathcal{F}(m,r_{\max})$ ordered by $\pi$}
  \State $k \gets k+1$
  \State $c \gets x_0 \oplus \bigoplus_{b\in S}2^b$
  \If{$c$ is prime}
    \Return $(c,\, k,\, |S|,\, |c-x_0|)$
  \EndIf
\EndFor
\end{algorithmic}
\end{algorithm}

The choice of ordering $\pi$ gives the following concrete strategies:

\begin{itemize}
  \item \textbf{LBF} (\emph{low-bit-first}): sort flip sets by
    $(\sum S,\max S)$, preferring lower bit indices.
  \item \textbf{HBF} (\emph{high-bit-first}): sort by
    $(-\sum S,-\max S)$.
  \item \textbf{RND} (\emph{uniform random}): shuffle flip sets randomly.
  \item \textbf{LRN} (\emph{learned bit-order}): sort by learned bit scores.
\end{itemize}

\subsection{Hybrid strategy}

The hybrid strategy interleaves SOI candidates and learned Hamming candidates,
deduplicating before testing.

\begin{algorithm}[H]
\caption{Hybrid standard + learned search}
\begin{algorithmic}[1]
\State $x_0 \gets x \,|\, 1$, $\mathrm{seen}\gets \emptyset$, $k\gets 0$
\For{$i=0,1,2,\ldots$}
  \If{$i<L$} add $x_0+2i$ to candidates \EndIf
  \If{$i<|\mathcal{F}|$} add $x_0\oplus \mathrm{mask}(\mathcal{F}_i)$ to candidates \EndIf
  \For{each candidate $c$ not in $\mathrm{seen}$}
    \State $\mathrm{seen}\gets \mathrm{seen}\cup\{c\}$, $k\gets k+1$
    \If{$c$ is prime}
      \Return $(c,k,\dH(x_0,c),|c-x_0|)$
    \EndIf
  \EndFor
\EndFor
\end{algorithmic}
\end{algorithm}

$L$ is the odd-increment limit, set to 128 in the reported experiments.

\subsection{Learned bit-order policy}
\label{sec:learned}

We learn per-bit usefulness scores from data. Given training dimension
$m_{\mathrm{tr}}$ and $N_{\mathrm{tr}}$ random odd starting points:

\begin{enumerate}
  \item Run LBF on each starting point and record the successful flip set $S^*$.
  \item For each bit $b\in S^*$, increment $\mathrm{score}[b]$.
  \item Bits with zero count receive a fallback score equal to
    $0.25$ times the minimum positive score.
  \item Bit~0 is forced to score 0 and is excluded from odd-subcube search.
\end{enumerate}

For test dimension $m_{\mathrm{te}}>m_{\mathrm{tr}}$, learned scores are
transferred by retaining the scores for old bit positions and assigning the
fallback score to new higher bit positions.

%% ---------------------------------------------------------------
\section{Heuristic Analysis}
\label{sec:theory}
%% ---------------------------------------------------------------

This section provides heuristic estimates and informal arguments. No new
theorems are proved.

\subsection{Expected checks for SOI}

The prime number theorem implies that the average gap between consecutive
primes near $N$ is approximately $\ln N$. Restricting to odd candidates halves
this, giving:
\[
  \E[\text{checks for SOI}]
  \approx
  \frac{\ln(2^m)}{2}
  =
  \frac{m\ln 2}{2}.
\]
For $m=32$, this gives:
\[
  \frac{32\ln 2}{2}\approx 11.1.
\]

\subsection{Hamming ball size and prime density}

The expected number of primes within Hamming radius $r$ of a random odd
$m$-bit integer is approximately:
\[
  \E[|P_m \cap \mathcal{B}(x_0,r)|]
  \approx
  B(m-1,r)\cdot \frac{2}{m\ln 2},
\]
where the factor of 2 accounts for restricting to odd candidates.

For $m=32$ and $r=4$:
\[
  B(31,4)
  =
  1+31+465+4495+31465
  =
  36457,
\]
and:
\[
  \frac{2}{32\ln 2}\approx 0.090.
\]
So the radius-4 odd-subcube ball contains many primes in expectation. The
difficulty is not whether a prime exists nearby in Hamming distance, but how
many ordered candidate checks are needed to reach one.

\subsection{Why SOI dominates average checks}

SOI tests candidates in increasing arithmetic order and directly exploits the
local prime density on the number line. Hamming search visits candidates in
bit-flip order, which does not align with arithmetic locality.

\begin{observation}
A Hamming flip of bit $k$ changes $x$ by $\pm 2^k$, which may be a very large
arithmetic step. Hamming search therefore optimizes bit locality rather than
number-line locality.
\end{observation}

\subsection{Why Hamming search wins on many individual inputs}

\begin{observation}
If a prime occurs early in the Hamming ordering of $x_0$, then Hamming search
can beat SOI for that starting point even when SOI has better average behavior.
\end{observation}

This accounts for the empirical fact that LBF wins or ties SOI on a large
fraction of individual inputs, despite losing on average candidate checks.

%% ---------------------------------------------------------------
\section{Experiments}
\label{sec:experiments}
%% ---------------------------------------------------------------

\subsection{Setup}

All primality tests use \texttt{sympy.isprime}. Results are cached with
\texttt{functools.lru\_cache}. The goal is not to benchmark primality-testing
algorithms themselves, but to compare candidate-generation policies. All
strategies in a given experiment use the same primality-testing backend.

Unless otherwise stated, the reported experiments use:

\begin{itemize}
  \item $m_{\mathrm{tr}}=20$
  \item $m_{\mathrm{te}}\in\{24,28,32\}$
  \item $N_{\mathrm{tr}}=5000$ training samples
  \item $N_{\mathrm{te}}=10000$ test samples per dimension
  \item $r_{\max}=4$
  \item seed $=42$
\end{itemize}

\subsection{Hamming distance distribution}

Table~\ref{tab:dist14} shows the full scan for $m=14$.

\begin{table}[h]
\centering
\caption{Hamming distance from each integer in $Q_{14}$ to its nearest prime.
Full scan, $2^{14}=16384$ integers.}
\label{tab:dist14}
\begin{tabular}{rrrr}
\toprule
$D_{14}(x)$ & Count & \% & Cumulative \% \\
\midrule
0 & 1900 & 11.6 & 11.6 \\
1 & 7933 & 48.4 & 60.0 \\
2 & 6286 & 38.4 & 98.4 \\
3 & 265  & 1.6  & 100.0 \\
$\geq 4$ & 0 & 0.0 & 100.0 \\
\bottomrule
\end{tabular}
\end{table}

This full scan shows that every integer in $Q_{14}$ is within Hamming distance
3 of a prime, and 98.4\% are within distance 2.

\subsection{Average candidate checks}

Table~\ref{tab:checks} reports average candidate checks by strategy.

\begin{table}[h]
\centering
\caption{Average candidate checks by strategy, $m \in \{24,28,32\}$,
$N_{\mathrm{te}}=10000$, $r_{\max}=4$.
SOI = standard odd-increment, LBF = low-bit-first odd-subcube search,
LRN = learned bit-order, HBF = high-bit-first, RND = uniform random.
Lower is better.}
\label{tab:checks}
\begin{tabular}{lrrr}
\toprule
Strategy & $m=24$ & $m=28$ & $m=32$ \\
\midrule
SOI            & \textbf{7.49}  & \textbf{8.73}  & \textbf{10.17} \\
Hybrid         & 7.70  & 9.15  & 10.58 \\
HBF            & 8.35  & 9.76  & 11.22 \\
LRN            & 8.37  & 9.96  & 11.56 \\
LBF            & 8.38  & 9.95  & 11.53 \\
RND            & 8.42  & 9.91  & 11.62 \\
\bottomrule
\end{tabular}
\end{table}

SOI has the fewest average checks at all tested dimensions. Although HBF has
competitive average checks in this table, it does so by testing candidates that
are often arithmetically very far from the starting point. Thus, candidate
checks alone are not a sufficient evaluation metric. Section~\ref{sec:pareto}
evaluates the strategies jointly by checks, Hamming distance, and arithmetic
distance.

\subsection{Per-instance wins}

Table~\ref{tab:wins} shows the pairwise win rate of LBF against SOI.

\begin{table}[h]
\centering
\caption{Per-instance win rate of LBF vs. SOI.
For each starting point $x$, we compare the number of candidate checks used by
each strategy.}
\label{tab:wins}
\begin{tabular}{lrrr}
\toprule
$m$ & LBF wins & SOI wins & Ties \\
\midrule
24 & 37.1\% & 36.6\% & 26.3\% \\
28 & 37.7\% & 39.0\% & 23.3\% \\
32 & 39.2\% & 40.1\% & 20.7\% \\
\bottomrule
\end{tabular}
\end{table}

Thus, LBF wins or ties SOI on 63.4\% ($m=24$), 60.3\% ($m=28$), and 59.9\%
($m=32$) of individual starting points, despite having a higher average check
count. This shows that SOI's average advantage does not imply uniform
per-instance dominance.

\subsection{Distance comparison}

Table~\ref{tab:distances} compares the average Hamming and arithmetic distances
to the found prime for $m=32$.

\begin{table}[h]
\centering
\caption{Average Hamming and arithmetic distance to the found prime, $m=32$.
Values are computed from the odd-subcube experiment with $N_{\mathrm{te}}=10000$
and $r_{\max}=4$.}
\label{tab:distances}
\begin{tabular}{lrr}
\toprule
Strategy & Avg $\dH(x_0,p)$ & Avg $|p-x_0|$ \\
\midrule
SOI    & 2.81 & \textbf{18.34} \\
Hybrid & 1.90 & 665{,}588.18 \\
HBF    & 0.98 & 486{,}532{,}672.18 \\
LBF    & \textbf{0.98} & 24{,}771{,}857.20 \\
LRN    & 0.98 & 25{,}453{,}898.40 \\
RND    & 0.98 & 86{,}389{,}309.05 \\
\bottomrule
\end{tabular}
\end{table}

Odd-subcube Hamming strategies find primes at roughly one bit flip from the
starting point on average. SOI, by contrast, finds primes at an average Hamming
distance of 2.81 bits for $m=32$, but with much smaller arithmetic distance.
This highlights the core tradeoff: SOI is arithmetic-local, while Hamming
search is bit-local.

%% ---------------------------------------------------------------
\section{Pareto Analysis}
\label{sec:pareto}
%% ---------------------------------------------------------------

\subsection{Pareto dominance}

\begin{definition}
Strategy $A$ \emph{Pareto-dominates} strategy $B$ on metrics $(f_1,f_2)$,
both to be minimized, if $f_1(A)\leq f_1(B)$ and $f_2(A)\leq f_2(B)$ with at
least one strict inequality. The \emph{Pareto frontier} is the set of
non-dominated strategies.
\end{definition}

\subsection{Checks vs. Hamming distance}

In the plane:
\[
  (\text{average checks},\ \text{average Hamming distance}),
\]
SOI and odd-subcube Hamming methods represent different tradeoffs. SOI has
the lowest average checks but larger Hamming distance. LBF, LRN, and related
Hamming strategies have slightly larger check counts but much smaller Hamming
distance.

This means neither family uniformly dominates when Hamming distance is included
as an objective.

\subsection{Checks vs. arithmetic distance}

In the plane:
\[
  (\text{average checks},\ \log_2(1+\text{average arithmetic distance})),
\]
SOI dominates most Hamming strategies because it is designed to find the next
prime on the number line. This confirms that SOI is the appropriate strategy
when arithmetic locality is the primary objective.

\subsection{Multi-objective score}

We define a composite score:
\[
  \text{score}
  =
  \overline{k}
  +
  \lambda_H\cdot \overline{\dH}
  +
  \lambda_A\cdot \log_2(1+\overline{|p-x_0|}),
\]
where $\overline{\cdot}$ denotes the average over all starting points.

The best strategy depends on the weights. When $\lambda_A$ is large or
candidate checks are weighted heavily, SOI is favored. When $\lambda_H$ is
large and arithmetic distance is weakly penalized, odd-subcube Hamming
strategies become competitive or preferred. This confirms that the strategies
are best understood as optimizing different geometries rather than as direct
replacements for one another.

%% ---------------------------------------------------------------
\section{Distribution Geometry}
\label{sec:distribution}
%% ---------------------------------------------------------------

This section asks not how to \emph{find} a prime in the hypercube, but how
primes are \emph{distributed} within it.  We compare the real prime set
$P_m \subset Q_m^{\mathrm{odd}}$ against a null model: a uniformly random
subset of $Q_m^{\mathrm{odd}}$ with the same empirical density $\delta_m$.

\subsection{Hamming-ball discrepancy}
\label{sec:discrepancy}

For a center $x$ and radius $r$, define the prime-count discrepancy:
\[
  \Delta(x,r) = |P_m \cap \mathcal{B}(x,r)| - \delta_m |\mathcal{B}(x,r)|,
\]
and the normalized z-score:
\[
  Z(x,r) = \frac{\Delta(x,r)}{\sqrt{\delta_m(1-\delta_m)|\mathcal{B}(x,r)|}}.
\]
The denominator is the standard deviation of prime count if primes were an
independent Bernoulli($\delta_m$) subset.

\begin{observation}
Under a Bernoulli random-subset model, $Z(x,r)$ has mean 0 and variance 1.
If primes are pseudorandom at the Hamming-ball scale, the empirical distribution
of $Z(x,r)$ over many centers $x$ should be close to $N(0,1)$.
\end{observation}

Running the discrepancy experiment at $m \in \{16,18,20\}$, radii
$r\in\{1,2,3,4\}$, and 2{,}000 random centers per $(m,r)$:

\begin{itemize}
  \item The mean z-score is within $0.05$ of zero for all tested $(m,r)$.
  \item Standard deviation of $Z$ is close to 1.0 for $r\geq 2$.
  \item Excess kurtosis is near zero (i.e., tail weight consistent with
    Gaussian), particularly at larger radii where ball sizes are large.
  \item The z-score distribution for real primes is statistically
    indistinguishable from that of a random baseline set of the same density.
\end{itemize}

\begin{remark}
At small radii ($r=1$), ball sizes are $m$ (the number of free bits), which
can be as small as 15 for $m=16$. With such small samples, the normal
approximation is weaker and variance estimates have high uncertainty.  For
$r\geq 2$, ball sizes exceed 100 and the Gaussian picture stabilizes.
\end{remark}

\subsection{Walsh--Fourier analysis}
\label{sec:fourier}

Define the centered prime indicator on $Q_m^{\mathrm{odd}}$:
\[
  g(x) = \mathbf{1}[x \in P_m] - \delta_m.
\]
Its Walsh--Fourier expansion is:
\[
  g(x) = \sum_{S \subseteq \{1,\ldots,m-1\}} \hat{g}(S)\,\chi_S(x),
\]
where $\chi_S(x) = (-1)^{\sum_{b\in S}\mathrm{bit}_b(x)}$ is the parity
character for bit subset $S$, and:
\[
  \hat{g}(S) = \frac{1}{|Q_m^{\mathrm{odd}}|} \sum_x g(x)\,\chi_S(x).
\]
The degree of a coefficient is $|S|$.  Large $|\hat{g}(S)|$ for small $|S|$
would indicate that primality correlates with the parity of few bits.

Running the analysis at $m\in\{12,14,16\}$ with maximum degree 3, and comparing
each coefficient against 30 random baseline sets of the same density:

\begin{itemize}
  \item All degree-1 and degree-2 coefficients are small
    ($|\hat{g}(S)|\lesssim 10^{-3}$), consistent with the Prime Number Theorem
    for arithmetic progressions.
  \item No low-degree coefficient exceeds the 95th percentile of the random
    baseline distribution in a statistically consistent manner.
  \item By Parseval's identity, the fraction of variance explained by
    degree-$\leq 3$ coefficients is below 5\% for $m\geq 14$ and shrinks
    as $m$ grows.
\end{itemize}

These findings are consistent with the view that primality is a high-complexity
Boolean function: most of its structure lives at high Fourier degree.

\subsection{Hamming-1 edge counts}
\label{sec:edge}

Define the prime-prime Hamming-1 edge count:
\[
  E_1 = |\{(p,q) : p,q \in P_m,\; \dH(p,q)=1\}|.
\]
Under the Bernoulli null model, the expected value is:
\[
  \mathbb{E}[E_1] \approx \delta_m^2 \cdot (m-1)\cdot 2^{m-2},
\]
since there are $(m-1)\cdot 2^{m-2}$ Hamming-1 edges in $Q_m^{\mathrm{odd}}$
and each independently has probability $\delta_m^2$ of connecting two primes.
The ratio $\rho = E_1 / \mathbb{E}[E_1]$ measures clustering ($\rho>1$) or
repulsion ($\rho<1$).

Table~\ref{tab:edges} reports results for $m=10$--$22$.

\begin{table}[h]
\centering
\caption{Prime-prime Hamming-1 edge count vs.\ random-model expectation.
$\rho$ = observed/expected. Baseline 95\% CI computed over 20 random seeds
of the same-density null set. The prime ratio is stable at $\approx 0.664$
across seven octaves of $m$.}
\label{tab:edges}
\begin{tabular}{rrrrrr}
\toprule
$m$ & Primes & $\rho$ & Baseline 95\% CI & $|\sigma|$ \\
\midrule
10 &     171 & 0.665 & $[0.921,\,1.069]$ &   8.7 \\
12 &     563 & 0.693 & $[0.953,\,1.047]$ &  12.7 \\
14 &   1{,}899 & 0.675 & $[0.969,\,1.026]$ &  22.4 \\
16 &   6{,}541 & 0.663 & $[0.983,\,1.017]$ &  38.0 \\
18 &  22{,}999 & 0.662 & $[0.991,\,1.006]$ &  87.9 \\
20 &  82{,}024 & 0.662 & $[0.995,\,1.004]$ & 155.4 \\
22 & 295{,}946 & 0.664 & $[0.996,\,1.002]$ & 221.5 \\
\bottomrule
\end{tabular}
\end{table}

\begin{observation}
The prime-prime Hamming-1 edge count is consistently $\approx 33\%$ below the
random-model expectation, with $\rho\approx 0.67$ stable across
$m=10$--$22$.  This is many standard deviations from the random baseline.
\end{observation}

\begin{observation}
Decomposing $E_1$ by bit position $j$ (where a bit-$j$ edge connects $x$ to
$x\oplus 2^j$), the ratio $\rho_j = E_1^{(j)}/\mathbb{E}[E_1^{(j)}]$ is
approximately \emph{uniform} across all $j\in\{1,\ldots,m-1\}$, deviating
by at most 5\% across bit positions at each $m$.
\end{observation}

\paragraph{Interpretation.}
The ratio $\rho\approx 0.67$ is consistent with sieve repulsion. For the bit-1
edge (step $\pm 2$, corresponding to twin primes), the Hardy--Littlewood
constant $C_2\approx 0.66$ predicts that twin prime pairs occur at a fraction
$C_2$ of the naive density-squared rate. The same modular-arithmetic obstruction
applies to all other step sizes $2^j$: given that $p$ is prime, small-prime
divisibility considerations reduce the probability that $p\pm 2^j$ is also
prime. The near-uniformity across bit positions reflects that the dominant
sieve correction (modulo 3) is independent of the step size.

\paragraph{Contrast with search geometry.}
In the search experiments, low bit positions are preferred: low-bit-first
ordering works because small bit flips correspond to small arithmetic changes,
where prime density on the number line is relevant. In the edge-count
experiment, there is \emph{no} bit-position gradient: all directions show the
same $\approx 33\%$ deficit. This contrast is a key finding:
\emph{search-path bias and prime-prime edge geometry are different phenomena
arising from different mechanisms.}

%% ---------------------------------------------------------------
\section{Related Work}
\label{sec:related}
%% ---------------------------------------------------------------

\paragraph{Prime gaps.}
The distribution of gaps between consecutive primes is well-studied
\citep{cramer1936, granville1995, maynard2015}. Our work is complementary:
we study not arithmetic gaps but Hamming distances, which encode a different
notion of proximity.

\paragraph{The binary hypercube.}
The hypercube is a classical object in combinatorics and coding theory
\citep{macwilliams1977, knuth2011}. Boolean functions on the hypercube have
been extensively analyzed \citep{odonnell2014}. The distribution of specific
number-theoretic sets, such as primes, as subsets of the hypercube appears less
standard as a unified search-geometry framework.

\paragraph{Primality testing.}
Miller--Rabin \citep{miller1976, rabin1980} and AKS \citep{agrawal2004} are
standard primality-testing references. This work does not study primality
testing itself; it studies candidate-generation policies.

\paragraph{Boolean Fourier analysis.}
The Walsh--Fourier analysis of Boolean functions on $\{0,1\}^m$ is a standard
tool in theoretical computer science \citep{odonnell2014}. Applying it to
number-theoretic indicator functions such as the prime indicator is less
standard; our experiments show that the prime indicator has small low-degree
coefficients, consistent with prime pseudorandomness.

\paragraph{Hardy--Littlewood conjectures and sieve theory.}
The constant $C_2\approx 0.66$ appearing in our edge-count ratio $\rho$
is the same constant from the Hardy--Littlewood twin prime conjecture
\citep{cramér1936, granville1995}.  The sieve interpretation of $\rho<1$ is
standard in analytic number theory, but the observation that this ratio is
uniform across all Hamming-1 directions (all step sizes $2^j$) is a
combinatorial-geometric formulation that does not appear in classical sources.

\paragraph{Learned heuristics for mathematical search.}
Machine-learning approaches to mathematical discovery and heuristic search
have been explored in recent work such as \citet{davies2021}. Our learned
bit-order policy is much simpler: it is a count-based heuristic rather than a
neural model.

%% ---------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}
%% ---------------------------------------------------------------

We embedded prime numbers in the odd binary hypercube and studied two
complementary questions: how does Hamming-ball prime search compare with
standard arithmetic search, and do primes behave like a random subset of the
odd hypercube?

The headline finding is a clean contrast:

\begin{center}
\emph{Primes look random inside Hamming balls, but repel along hypercube edges.}
\end{center}

More precisely, the empirical findings are:

\paragraph{Search geometry.}
\begin{enumerate}
  \item SOI achieves the lowest average candidate-check count, but LBF wins or
    ties SOI on 58--63\% of individual starting points.

  \item The two search families are Pareto-incomparable: SOI minimizes
    arithmetic distance, Hamming search minimizes Hamming distance.  Neither
    dominates when both objectives are evaluated.

  \item Bit-position bias in search paths (low bits preferred) reflects
    arithmetic locality of the number line, not prime-prime Hamming geometry.
\end{enumerate}

\paragraph{Distribution geometry.}
\begin{enumerate}
  \setcounter{enumi}{3}
  \item Hamming-ball prime counts fluctuate with z-scores close to $N(0,1)$,
    indistinguishable from a random subset of the same density.

  \item Low-degree Walsh--Fourier coefficients of the prime indicator are
    consistent with random-baseline magnitudes; primes do not cluster by
    simple parity patterns.

  \item Prime-prime Hamming-1 edge counts are $\approx 33\%$ below the
    random-model expectation ($\rho\approx 0.67$), uniform across all bit
    positions and stable across $m=10$--$22$.  This is consistent with
    Hardy--Littlewood sieve corrections.

  \item Search-path bias and prime-prime edge repulsion are distinct
    phenomena.  The first is bit-position-dependent; the second is not.
    This distinction is not obvious at the outset.
\end{enumerate}

\subsection{What this framework provides}

The overall contribution is an experimental hypercube-geometry framework for
separating where primes look random and where arithmetic structure becomes
visible.  The better framing is not ``Hamming search vs. arithmetic search,''
but rather:

\begin{quote}
\emph{When does prime distribution in the hypercube reveal arithmetic
structure, and at what scale?}
\end{quote}

The answer from these experiments: pseudorandom at the ball and Fourier scale,
non-random at the single-edge scale.

\subsection{Limitations}

This work is empirical. No asymptotic claims are proved.  Experiments cover
moderate bit sizes ($m\leq 32$ for search, $m\leq 22$ for edge counts, $m\leq
20$ for Fourier).  The Hamming strategies tested are simple orderings of
bit-flip masks; residue-aware policies may change the search tradeoffs.  The
edge-count ratio $\rho\approx 0.67$ is consistent with Hardy--Littlewood
constants but this connection is not proved rigorously here.

\subsection{Open questions}

\begin{enumerate}[label=(\roman*)]
  \item \textbf{Asymptotic edge ratio.} Does $\rho\to C_2$ as $m\to\infty$,
    where $C_2\approx 0.66$ is the Hardy--Littlewood twin-prime constant?  A
    rigorous connection would require relating bit-1 Hamming-1 edges to twin
    prime pairs and applying known sieve estimates.

  \item \textbf{Hamming-distance-2 sieve repulsion.} The bit-1 edge deficit
    reflects a sieve correction for step~$\pm 2$.  Do prime pairs at Hamming
    distance 2 (two bit flips) show a corresponding correction, and how does
    it relate to known constellations such as $\{p, p+2, p+6\}$?

  \item \textbf{Uniformity across bit positions.} Is the near-uniform ratio
    $\rho_j\approx 0.67$ across all $j$ provably implied by standard sieve
    theory, or could a careful residue argument produce a gradient?

  \item \textbf{Hypercube sieve.} Can modular structure be used to prescreen
    Hamming candidates, eliminating those divisible by 3, 5, 7 before the
    primality test?

  \item \textbf{Larger search dimensions.} Do the Pareto structure and win
    rates hold at $m\in\{40,48,64\}$?

  \item \textbf{Fourier at larger $m$.} Can randomized or sketching-based
    Fourier methods estimate low-degree coefficients at $m\geq 24$ without
    enumerating all $2^{m-1}$ vertices?
\end{enumerate}

\subsection*{Code availability}

All code, data generation scripts, analysis, and blog chapter drafts are
available at:
\[
  \text{\url{https://github.com/pojhafb/prime-hypercube-search}}
\]

%% ---------------------------------------------------------------
\appendix
%% ---------------------------------------------------------------

\section{Flip-set construction}
\label{app:flipsets}

For $m$ bits and maximum radius $r_{\max}$, the full list of flip sets in the
odd subcube is:
\[
  \mathcal{F}(m,r_{\max})
  =
  \bigcup_{r=0}^{r_{\max}}
  \binom{\{1,\ldots,m-1\}}{r},
\]
where $\binom{S}{r}$ denotes all $r$-element subsets of $S$.

The total number of flip sets is $B(m-1,r_{\max})$.

\begin{center}
\begin{tabular}{rrrrrr}
\toprule
$m$ & $r=0$ & $r=1$ & $r=2$ & $r=3$ & $r=4$ total \\
\midrule
16 & 1 & 15 & 105 & 455 & 1941 \\
32 & 1 & 31 & 465 & 4495 & 36457 \\
64 & 1 & 63 & 1953 & 39711 & 637393 \\
\bottomrule
\end{tabular}
\end{center}

At $m=64$, the radius-4 odd-subcube flip-set list contains 637,393 candidates.
Caching per $(m,\text{strategy})$ reduces this to a one-time cost per experiment.

\section{Primality testing}
\label{app:primality}

We use \texttt{sympy.isprime}, which combines trial division and primality tests
such as Miller--Rabin and Baillie--PSW depending on the size of the input.
Results are cached via \texttt{functools.lru\_cache} with capacity
$2\times 10^6$.

The goal of this work is not to benchmark primality-testing algorithms
themselves, but to compare candidate-generation policies. All strategies in a
given experiment use the same primality-testing backend.

%% ---------------------------------------------------------------
\begin{thebibliography}{99}

\bibitem[Agrawal et~al.(2004)]{agrawal2004}
Agrawal, M., Kayal, N., and Saxena, N. (2004).
\newblock PRIMES is in P.
\newblock \emph{Annals of Mathematics}, 160(2), 781--793.

\bibitem[Cram\'er(1936)]{cramer1936}
Cram\'er, H. (1936).
\newblock On the order of magnitude of the difference between consecutive prime numbers.
\newblock \emph{Acta Arithmetica}, 2, 23--46.

\bibitem[Davies et~al.(2021)]{davies2021}
Davies, A., Veli\v{c}kovi\'c, P., Buesing, L., et~al. (2021).
\newblock Advancing mathematics by guiding human intuition with AI.
\newblock \emph{Nature}, 600, 70--74.

\bibitem[de~la Vall\'ee~Poussin(1896)]{delavalleepoussin1896}
de~la Vall\'ee~Poussin, C.~J. (1896).
\newblock Recherches analytiques sur la th\'eorie des nombres premiers.
\newblock \emph{Annales de la Soci\'et\'e Scientifique de Bruxelles}, 20, 183--256.

\bibitem[Granville(1995)]{granville1995}
Granville, A. (1995).
\newblock Harald Cram\'er and the distribution of prime numbers.
\newblock \emph{Scandinavian Actuarial Journal}, 1995(1), 12--28.

\bibitem[Hadamard(1896)]{hadamard1896}
Hadamard, J. (1896).
\newblock Sur la distribution des z\'eros de la fonction $\zeta(s)$ et ses
  cons\'equences arithm\'etiques.
\newblock \emph{Bulletin de la Soci\'et\'e Math\'ematique de France}, 24, 199--220.

\bibitem[Knuth(2011)]{knuth2011}
Knuth, D.~E. (2011).
\newblock \emph{The Art of Computer Programming, Volume~4A: Combinatorial
  Algorithms, Part~1}.
\newblock Addison-Wesley.

\bibitem[MacWilliams and Sloane(1977)]{macwilliams1977}
MacWilliams, F.~J. and Sloane, N.~J.~A. (1977).
\newblock \emph{The Theory of Error-Correcting Codes}.
\newblock North-Holland.

\bibitem[Maynard(2015)]{maynard2015}
Maynard, J. (2015).
\newblock Small gaps between primes.
\newblock \emph{Annals of Mathematics}, 181(1), 383--413.

\bibitem[Menezes et~al.(1996)]{menezes1996}
Menezes, A.~J., van Oorschot, P.~C., and Vanstone, S.~A. (1996).
\newblock \emph{Handbook of Applied Cryptography}.
\newblock CRC Press.

\bibitem[Miller(1976)]{miller1976}
Miller, G.~L. (1976).
\newblock Riemann's hypothesis and tests for primality.
\newblock \emph{Journal of Computer and System Sciences}, 13(3), 300--317.

\bibitem[O'Donnell(2014)]{odonnell2014}
O'Donnell, R. (2014).
\newblock \emph{Analysis of Boolean Functions}.
\newblock Cambridge University Press.

\bibitem[Rabin(1980)]{rabin1980}
Rabin, M.~O. (1980).
\newblock Probabilistic algorithm for testing primality.
\newblock \emph{Journal of Number Theory}, 12(1), 128--138.

\end{thebibliography}

%% ---------------------------------------------------------------
\end{document}
